---
title: "Lake Water Quality Analysis"
author: "Sam Leuthold"
date: "02"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(lubridate) #For dealing with date and time
library(knitr)
library(kableExtra)
```

# Part 1

## Set-up
 
```{r data-read}
# Lagos download script
#LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

lagos <- lagosne_load()

lake_centers <- lagos$locus

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

mapview(subset_spatial)

states <- us_states()

minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

minnesota_lakes <- spatial_lakes[minnesota,]

minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')

```



# In-Class work


## 1) Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}

il.ia.states <- states %>% 
                  filter(name %in% c("Iowa", "Illinois")) %>%
                  st_transform(2163)

mapview(il.ia.states)

```



## 2) Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

```{r}

il.ia.lakes <-  spatial_lakes[il.ia.states,]
                  
nrow(il.ia.lakes)

nrow(minnesota_lakes)

```


## 3) What is the distribution of lake size in Iowa vs. Minnesota?

- Here I want to see a histogram plot with lake size on x-axis and frequency on y axis (check out geom_histogram)

```{r}

combined.lake.data <- il.ia.states %>%
                        filter(name == "Iowa") %>%
                        spatial_lakes[.,] %>%
                        bind_rows(minnesota_lakes) %>%
                        mutate(State_Name = ifelse(state_zoneid == "State_13", 
                                                   "Minnesota",
                                                   "Iowa"))



ggplot(combined.lake.data) +
  geom_histogram(aes(x = lake_area_ha),
                 color = "black",) +
  facet_grid(.~State_Name) +
  scale_x_log10() +
  theme_light()

```


## 4) Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r}
mapview(il.ia.lakes,
        zcol = 'lake_area_ha')

```


## 5) What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

Remote sensing data provides another means for estimating lake size across states. An added advantage to this method is the ability to analyze the shifts in lake size temporally, which can be especially relevant in systems that provide drinking water to downstream communities. 


# Part 2

## Set-up (again)

```{r data-read2}

lagos <- lagosne_load()

lake_centers <- lagos$locus

spatial_lakes <- st_as_sf(lake_centers,
                          coords=c('nhd_long',
                                   'nhd_lat'),
                          crs=4326)

nutr <- lagos$epi_nutr

clarity_only <- nutr %>%
  select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)


spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')

mean_values_200 <- chla_secchi_200 %>%
                    group_by(lagoslakeid) %>%
                    summarize(mean_chl = mean(chla,na.rm=T),
                              mean_secchi=mean(secchi,na.rm=T)) %>%
                    filter(!is.na(mean_chl),
                           !is.na(mean_secchi)) %>%
                    mutate(log10_mean_chl = log10(mean_chl))

mean_spatial <- inner_join(spatial_lakes,mean_values_200,
                           by='lagoslakeid') 
```


# Class work

## 1) What is the correlation between Secchi Disk Depth and Chlorophyll a for sites with at least 200 observations?

- Here, I just want a plot of chla vs secchi for all sites 

```{r, message=FALSE}

ggplot(mean_values_200,
       aes(x = mean_secchi,
           y = log(mean_chl))) +
  geom_smooth(method = "lm",
              se = F,
              color = "grey50",
              lty = 2) +
  annotate(geom = "text",
           x = 7, 
           y = 3.5,
           label = paste0("r = ", round(cor(mean_values_200$mean_secchi, 
                                            mean_values_200$log10_mean_chl), 3))) +
  xlab("Average Secchi Disk Depth") +
  ylab("Log(Chlorophyll A)") +
  geom_point(size = 2) +
  theme_light()


```


## Why might this be the case? 

Increased chlorophyll A is an indicator of increased phytoplankton growth, which can obscure water clarity and reduce Secchi disk depth.

## 2) What states have the most data? 

### 2a) First you will need to make a lagos spatial dataset that has the total number of counts per site.

```{r}

nutrient.data <- lagos$epi_nutr

counts.data <- nutrient.data %>%
                group_by(lagoslakeid) %>%
                mutate(count = n()) %>% 
                inner_join(spatial_lakes, .,
                           by='lagoslakeid') 


```


### 2b) Second, you will need to join this point dataset to the us_boundaries 
data. 

```{r}

states <- us_states()

counts.data <- st_join(counts.data, states)

```


### 2c) Then you will want to group by state and sum all the observations in that state and arrange that data from most to least toatl observations per state. 

```{r}

lake.counts <- aggregate(count ~ state_name, data = counts.data, FUN = sum)

lake.counts <- arrange(lake.counts, desc(count))

lake.counts %>%
  kbl(col.names = c("State", "Number of Obs.")) %>%
  kable_minimal()
```

##3 Is there a spatial pattern in Secchi disk depth for lakes with at least 200 observations?

Yes, there is a general spatial pattern in the Secchi disk depth. Of the lakes with at least 200 observations, lakes in the major agricultural regions (i.e., the central United States) tend to have a more shallow average Secchi disk depth value. In contrast, lakes in the upper eastern portion of the country have deeper Secchi disk depth values, possibly due to a decrease in the amount of phytoplankton growth facilitated by fertilizer runoff and grounwater nutrient loading. 

```{r}
mapview(mean_spatial,
        zcol = "mean_secchi")
```



